The mass dependence of halo alignments with large-scale structure

### 2017-05-05 draft.tex
Add files via upload

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[\textit{temporary} Slope of intrinsic alignment amplitude vs mass]{\textit{temporary} The mass dependence of halo alignments with large-scale structure}

@@

\section{Introduction}
\label{sec:intro}
When looking at the sky, assuming a homogeneous and isotropic universe, one would expect to see randomly distributed galaxies with arbitrary orientation, whereas correlation between galaxy shapes is actually observed; plenty of reasons can be exhibited to explain this, such as the effect of gravitational tidal fields from the surrounding dark matter haloes and the accretion of new material along favoured directions. This phenomenon is known as intrinsic alignment (sometimes referred to as IA; see \citet{Joachimietal2015} for a recent review): besides carrying information about galaxy formation, it contributes to the overall cosmic shear signal, and must therefore be taken into consideration in order not to bias the results from future surveys, such as LSST \citep{LSST2009} and Euclid \citep{Euclid2011}.

Measurements of the intrinsic alignment signal have been performed using data from observations \citep{Mandelbaumetal2006, Hirataetal2007, Okumuraetal2009} and from hydrodynamical simulations \citep{Codisetal2015, Velliscigetal2015, Chisarietal2015, Hilbertetal2016, Tennetietal2016}, suggesting that massive red galaxies point towards matter overdensities; as far as blue galaxies are concerned, on the other hand, no clear detection of intrinsic alignment has been already claimed \citep{Hirataetal2007, Mandelbaumetal2011}.

The dependence of the intrinsic alignment signal with mass has already been studied using simulation data \citep{Jing2002}
In this work we delve into the dependence of the amplitude of the intrinsic alignment signal on the mass of the halo: while for galaxies \citep{Joachimietal2011, Singhetal2015} and for clusters \citep{vanUitertJoachimi2017} an increasing trend with luminosity, or the corresponding mass, has already been identified, we analyse the trend for both galaxies and galaxy clusters using simulation and real data (Sect.~\ref{sec:data}). We perform a Bayesian analysis, described in Sect.~\ref{sec:method}, to establish the slope of the power-law model we assume to describe the IA amplitude signal, and show our results in Sect.~\ref{sec:resanddiscuss}.
\section{Data}
\label{sec:data}
\subsection{Simulations}
\label{subsec:sim}
In this work we consider haloes from two different simulations: 
\begin{enumerate}
\item the \textbf{Millennium Simulation (MS)}, first presented in \citet{Springeletal2005}, which uses $2160^3$ dark matter particles of mass $m_{\mathrm{P}}^{\mathrm{MS}} =1.2 \times 10^9 M_{\sun}$ enclosed in a 500 Mpc$/h$-side box to sample dark matter haloes and study their growth. In particular, we take into consideration 2 of its 64 snapshots, i.e.\ the one at $z = 0$ (snapshot $63$), which we use for all our analysis, and the one at $z \simeq 0.46$ (snapshot $49$), which is studied only as an aside. Dark matter haloes are identified as in \citet{Joachimietal2013a} and references therein, assuming that all galaxies share the three-dimensional shape of their host halo. All haloes have a minimum number of particles $N_{\mathrm{P}} = 100$.
\item The \textbf{Millennium-XXL Simulation (XXL)}, which samples $6720^3$ dark matter particles of mass $m_{\mathrm{P}}^{\mathrm{XXL}} = 8.456 \times 10^9 M_{\sun}$ confined in a cubic region of $3000$ Mpc$/h$ on a side \citep{Anguloetal2012}; in this case, we consider only one snapshot, at $z = 0$. Particles are selected using an ellipsoidal overdensity algorithm, as described in \citet{Despalietal2013} and \citet{Bonamigoetal2015}.
\end{enumerate} 
%the shape of the haloes is then derived as for the MS from the eigenvectors and eigenvalues of the mass tensor 
%\begin{equation}
%    \mathbf{M}_{\mu \nu} = m_P \sum_{i=1}^{N_P} \frac{r_{i, \mu} r_{i, \nu}}{M_{TOT}} = \frac{1}{N_P} \sum_{i=1}^{N_P} r_{i, \mu} r_{i, \nu}
%	\label{eq:mt}
%\end{equation}
%with $M_{TOT}$ the total mass of the object. Note that the mass tensor (Eq.~\ref{eq:mt}) and the inertia tensor (Eq.~\ref{eq:sit}) are different quantities \citep{Bettetal2007}, but the analysis on the two simulations yields pretty consistent results, as will be shown; also, for this catalogue no ``reduced'' tensor is taken into consideration.

@@

To compare our results with real data, we also take into consideration the analysis presented in \citet{vanUitertJoachimi2017}: in that work, the clusters contained in the redMaPPer catalogue \citep{Rykoffetal2014} version 6.3 were used to calculate the intrinsic alignment signal amplitude $A_{\mathrm{IA}}$, which was then studied as a function of the halo mass.

\subsection{Halo shapes}
\label{subsec:haloshapes}
The shape of each halo for both simulations is described by the eigenvalues and eigenvectors of the simple inertia tensor\footnote{MS and XXL use two different tensors to describe the shape, but our choice has no impact on the ellipticity of the haloes (see also \citet{Bettetal2007} for further details).}
\begin{equation}
    \mathbf{M}_{\mu \nu} \propto \sum_{i=1}^{N_{\mathrm{P}}} r_{i, \mu} r_{i, \nu} \ ,
	\label{eq:sit}
\end{equation}
where $N_{\mathrm{P}}$ is the total number of particles within the halo, and $\mathbf{r}_{i}$ is the vector that indicates the position of the $i-$th particle with respect to the centre of the halo, i.e.\ the location of the gravitational potential minimum. For the MS only, we also consider a reduced inertia tensor, which is defined as \citep{Pereiraetal2008}:
\begin{equation}
    \mathbf{M}_{\mu \nu}^{\mathrm{red}} \propto \sum _{i=1}^{N_{\mathrm{P}}} \frac{r_{i, \mu} r_{i, \nu}}{r_i^2} \ ,
	\label{eq:rit}
\end{equation}
with $r_i^2$ the square of the three-dimensional distance of the i-th particle from the centre of the halo; the reduced inertia tensor is more weighted towards the centre of the halo, and may yield a more reliable approximation of the shape of the galaxy \citep{Joachimietal2013b, Chisarietal2015}.

The eigenvectors define an ellipsoid, which we project onto one of the faces of the cube that hosts the simulation: the resulting ellipse is the shape of the galaxy, which is assumed to be the same as the shape of the host halo. We proceed as in \citet{Joachimietal2013a, Joachimietal2013b} to define the ellipticity of the galaxy, considering our objects as early-type galaxies: indicating the three unit eigenvectors as $\mathbf{s}_{\mu} = \big \{ s_{x, \mu}, s_{y, \mu}, s_{\mathbin{\|}, \mu}\big \}^\intercal$ and the absolute values of the semi-axes as $\omega_{\mu}$, $\mu \in \{1,2,3\}$, we define a symmetric tensor
\begin{equation}
    \mathbf{W}^{-1} = \sum _{\mu=1}^{3} \frac{s_{\perp, \mu} s^\intercal_{\perp, \mu}}{\omega_{\mu}^2} - \frac{\mathbf{\kappa}\mathbf{\kappa}^{\intercal}}{\alpha^2} \ ,
	\label{eq:symtensor}
\end{equation}
with $s_{\perp, \mu} = \big \{ s_{x, \mu}, s_{y, \mu}\big \}^\intercal$ the eigenvector projected along the line of sight, 
\begin{equation}
   \mathbf{\kappa} = \sum_{\mu = 1}^{3} \frac{s_{\mathbin{\|}, \mu} \mathbf{s}_{\perp, \mu}}{\omega_{\mu}^2} \ ,
	\label{eq:kappa}
\end{equation}
and 
\begin{equation}
    \alpha^2=\sum_{\mu = 1}^{3} \bigg( \frac{s_{\mathbin{\|}, \mu}}{\omega_{\mu}} \bigg)^2 \ .
	\label{eq:alpha}
\end{equation}
We then compute the two components of the ellipticity $\epsilon$ \citep{BartelmannSchneider2001}:
\begin{align}
    	\epsilon_{1} = \frac{W_{11} - W_{22}}{W_{11} + W_{22}+2\sqrt{\det \mathbf{W}}}\ ; \\ 
           \epsilon_{2} = \frac{2 W_{12}}{W_{11} + W_{22}+2\sqrt{\det \mathbf{W}}} \ .
\end{align}

@@

We describe the clustering signal with a simple model:
\begin{equation}
    w_{\mathrm{gg}} (R_{\mathrm{p}}, M)=b_g^2 (M)\ w_{\mathrm{\delta} \mathrm{\delta}}^{\mathrm{model}} (R_{\mathrm{p}}) - C_{\mathrm{IC}} \ , 
	\label{eq:wgg}
\end{equation}
with $w_{\mathrm{\delta} \mathrm{\delta}}^{\mathrm{model}} (R_{\mathrm{p}})$ a function in which we include the dependence on $R_{\mathrm{p}}$ \citep[equation 9]{vanUitertJoachimi2017}, and $C_{\mathrm{IC}}$ the integral constraint factor, which we estimate as in \citet[equation 8]{RocheEales1999} using the random pair counts. Again, we evaluate the previous expression at $R_{\mathrm{p}}^*$, obtaining:
\begin{equation}
    w_{\mathrm{gg}} (M)= w_{\mathrm{gg}} (R_{\mathrm{p}} = R_{\mathrm{p}}^*, M) \ .
	\label{eq:wgpmass}
\end{equation}

@@

To achieve this goal, we select the objects of the catalogues described in Sect.~\ref{subsec:sim} in $n = 5$ logarithmic mass bins, between $10^{11} M_{\sun}$ and $10^{14} M_{\sun}$ for the MS and between $10^{12.5} M_{\sun}$ and $10^{15} M_{\sun}$ for the XXL, we split them in $N = 3^3 = 27$ sub-boxes based on their positions inside the cube of the respective simulation, and calculate $w_{\mathrm{g+}}$ and $w_{\mathrm{gg}
}$ for each of the N sub-samples by replacing the integrals in Eq.~\ref{eq:wgphat} and \ref{eq:wgghat} with a sum over 20 line-of-sight bins, each $(\Pi_{\mathrm{max}} - (-\Pi_{\mathrm{max}}))/ 20 \ = 6 \ \mbox{Mpc}/h$ wide. We show the distribution of the mass\footnote{Check the threshold, and explain better that this is not $M_{\mathrm{200m}}$, but the original one (virial).} for the two catalogues in Fig.~\ref{fig:histo}. 
\begin{figure}
	\centerline{
	\includegraphics[scale = 0.5]{img/histo.eps}}
	\caption{Histogram showing the distribution of the mass of the haloes: the Millennium-XXL catalogue (black line) contains about 6 times more objects than the Millennium catalogue (blue line). Note that in our analysis the Millennium-XXL catalogue is complete only above $10^{13} M_{\sun}$.}
	\label{fig:histo}
\end{figure}

To measure $DR$ and $RR$ as in Eq.~\ref{eq:xigghat}, we generate random catalogues that contain objects uniformly distributed between the minimum and maximum value of the $x$, $y$ and $z$ coordinates of each sub-box. These catalogues normally are 3 times denser; in some cases, when a sub-box encloses very few objects, we switch to 10-times-denser random catalogues.

We then perform a posterior analysis over the data to retrieve the distributions of $A$ and $\beta$: according to Bayes' theorem, if $\boldsymbol{d}$ is the vector of the data and $\boldsymbol{p}$ the vector of the parameters, 
\begin{equation}
    P(\boldsymbol{p} | \boldsymbol{d}) \propto P(\boldsymbol{d} | \boldsymbol{p}) \ P(\boldsymbol{p}) \propto e^{-\frac{1}{2} \chi ^2} P(\boldsymbol{p}) \ ,
	\label{eq:bayes}
\end{equation}
with $P(\boldsymbol{p} | \boldsymbol{d})$ the posterior probability, $P(\boldsymbol{d} | \boldsymbol{p})$ the likelihood function, $P(\boldsymbol{p})$ the prior probability and $\chi ^2 = (\boldsymbol{d} - \boldsymbol{m})^T \mathbf{C}^{-1} (\boldsymbol{d} - \boldsymbol{m})$, with $\boldsymbol{m}$ the vector of the model and $\mathbf{C}^{-1}$ the precision matrix, the inverse of the covariance matrix $\mathbf{C}$. We assume uninformative flat priors in the fit with ranges $\log_{10} A \in [-2.3;-0.5]$ and $\beta \in [-0.50;0.75]$.

We estimate the covariance matrix from the data as in \citet{Tayloretal2013}:
\begin{equation}
     \mathbf{C}_{\mu \nu} = \frac{1}{N-1} \sum_{j = 1}^{N} (x_{j, \mu} - \overline{x}_{\mu})(x_{j, \nu} - \overline{x}_{\nu}) \ ,
	\label{eq:covariance}
\end{equation}
with $\mu, \nu \in \{1, \dotso, n\}, \overline{x}_{\mu} = \frac{1}{N} \sum_{m=1}^{N} x_{m, \mu},$ and $x_{m, \mu}= r_{\mathrm{g+}}(M)$ for each sub-box and each mass bin, as defined in Eq.~\ref{eq:rg+}. We then invert the covariance matrix and correct the bias on the inverse to obtain an unbiased estimate of the precision matrix, given by:
\begin{equation}
     \mathbf{C}^{-1}_{\mathrm{unbiased}} = \frac{N - n -2}{N-1} \  \mathbf{C}^{-1} \ ,
	\label{eq:precunbiased}
\end{equation}
where $N > n+2$ clearly holds. The results of the analysis are presented in Sect.~\ref{sec:resanddiscuss}.

@@

%We thank Mario Bonamigo for providing us with the Millennium-XXL catalogue snapshots, and Edo van Uitert for making the real data points available to us and for useful discussion about the correct mass units of measurement to use. BJ acknowledges support by an STFC Ernest Rutherford Fellowship, grant reference ST/J004421/1.

@@

%\bibliographystyle{mnras}
%\bibliography{example} % if your bibtex file is called example.bib

% Alternatively you could enter them by hand, like this:
% This method is tedious and prone to error if you have lots of references
\begin{thebibliography}{99}
\bibitem[\protect\citeauthoryear{Angulo et al.}{2012}]{Anguloetal2012}
Angulo R.~E., Springel V., White S.~D.~M., et al., 2012, \href{https://academic.oup.com/mnras/article-lookup/doi/10.1111/j.1365-2966.2012.21830.x}{MNRAS}, \href{http://adsabs.harvard.edu/abs/2012MNRAS.426.2046A}{426}, \href{http://adsabs.harvard.edu/abs/2012MNRAS.426.2046A}{2046}
\bibitem[\protect\citeauthoryear{Bartelmann \& Schneider}{2001}]{BartelmannSchneider2001}
Bartelmann M., Schneider P., 2001, \href{http://www.sciencedirect.com/science/article/pii/S037015730000082X}{Phys. Reports}, \href{http://adsabs.harvard.edu/abs/2001PhR...340..291B}{340}, \href{http://adsabs.harvard.edu/abs/2001PhR...340..291B}{291}
\bibitem[\protect\citeauthoryear{Bett et al.}{2007}]{Bettetal2007}
Bett P., Eke V., Frenk C.~.S., et al., 2007, \href{https://academic.oup.com/mnras/article-lookup/doi/10.1111/j.1365-2966.2007.11432.x}{MNRAS}, \href{http://adsabs.harvard.edu/abs/2007MNRAS.376..215B}{376}, \href{http://adsabs.harvard.edu/abs/2007MNRAS.376..215B}{215}
\bibitem[\protect\citeauthoryear{Bonamigo et al.}{2015}]{Bonamigoetal2015}

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Mass dependence of halo alignments]{The mass dependence of halo alignments with large-scale structure}

@@

\section{Introduction}
\label{sec:intro}
When looking at the sky, assuming a homogeneous and isotropic universe, one would expect to see randomly distributed galaxies with arbitrary orientation, whereas correlation between galaxy shapes is actually observed. Plenty of reasons can be exhibited to explain this, such as the effect of gravitational tidal fields from the surrounding dark matter haloes and the accretion of new material along favoured directions: this phenomenon is known as intrinsic alignment (sometimes referred to as IA; see \citet{TroxelIshak2015, Joachimietal2015, Kiesslingetal2015} for recent reviews). Besides carrying information about galaxy formation, IA contributes to the overall cosmic shear signal: \citet{Heavensetal2000} and \citet{CroftMetzler2000} first showed the non-negligible contamination of the weak lensing signal due to correlations in the intrinsic shapes of galaxies using simulated data, and many other works afterwards (\citet{Heymansetal2006, Sembolonietal2008}, just to mention two of them) confirmed that this effect must be taken into consideration in order not to bias the results from deep lensing surveys, such as LSST \citep{LSST2009} and Euclid \citep{Euclid2011}.

Measurements of the intrinsic alignment signal have been performed using data from hydrodynamical simulations \citep{Codisetal2015, Velliscigetal2015, Chisarietal2015, Hilbertetal2017, Tennetietal2016}, from N-body simulations \citep{Heavensetal2000, CroftMetzler2000, Jing2002, Kuhlenetal2007, Schneideretal2012}, as already mentioned, and considering observation data as well \citep{Mandelbaumetal2006, Hirataetal2007, Okumuraetal2009, Haoetal2011, vanUitertJoachimi2017}, claiming that massive red galaxies point towards matter overdensities, while blue galaxies do not show any clear hint of alignment \citep{Hirataetal2007, Mandelbaumetal2011}.

The dependence of the IA signal with redshift and mass has also been studied: \citet{Haoetal2011}, for example, studied a large sample of galaxy clusters from the SDSS DR7 finding a strong dependence of the alignment with redshift and an increasing trend with the mass of the brightest cluster galaxy. Moreover, \citet{Jing2002} used a $512^3$-particle simulation to get accurate formulae to fit the ellipticity correlation function over three orders of magnitude in mass, while an increasing trend of the IA amplitude with luminosity, or the corresponding mass, has been identified using real data, for galaxies \citep{Joachimietal2011, Singhetal2015} and for clusters \citep{vanUitertJoachimi2017}. In order to assume more stringent priors on the parameters of the models that will be adopted by future surveys, it is important to obtain accurate constraints on these dependences.

In this work, we delve into the dependence of the amplitude of the intrinsic alignment signal on the mass of the halo: after a brief theory overview (Sect.~\ref{sec:theory}), we analyse the trend for both galaxies and galaxy clusters using simulation and real data (Sect.~\ref{sec:data}), we perform a Bayesian analysis, described in Sect.~\ref{sec:method}, to establish the slope of the power-law model we assume to describe the IA amplitude signal, and show our results in Sect.~\ref{sec:resanddiscuss}.
\section{Theoretical background}
\label{sec:theory}
The physical picture of tidal interaction of a self-gravitating system that is in virial equilibrium with a velocity dispersion $\sigma$, such as an elliptical galaxy or a cluster of galaxies, would be a distortion of the system's gravitational potential through tidal gravitational forces. The particles of the system would remain in virial equilibrium and fill up the distorted potential along an isocontour of the gravitational potential, which would result in a change in the shape of the system. This shape should reflect the magnitude and the orientation of the tidal gravitational fields, and the magnitude of the change in shape should depend on how tightly the system is bound.

Gravitational tidal fields generated by the ambient large-scale structure would distort the gravitational potential, and we will work in the limit that the distortion is well-described by a second-order Taylor-expansion relative to the potential minimum,
\begin{equation}
\Phi\rightarrow\Phi+\frac{1}{2}\frac{\partial^2\Phi}{\partial r_i\partial r_j}r_ir_j
\end{equation}
with the distance $r_i$ from the minimum. Consequently, the density of particles would change according to
\begin{equation}
\rho\propto\exp\left(-\frac{\Phi}{\sigma^2}\right) \rightarrow 
\exp\left(-\frac{\Phi}{\sigma^2}\right)\left(1+\frac{1}{2\sigma^2}\frac{\partial^2\Phi}{\partial r_i\partial r_j}r_ir_j\right),
\end{equation}
and a measurement of the ellipticity through the second moments of $\rho$ would yield a proportionality to $R^2\partial^2\Phi/\sigma^2$, with $R$ the size of the object. The expansion is only valid in the limit of weak tidal fields, which is characterised by $R^2\partial^2\Phi/\sigma^2\ll 1$.

The variance of tidal shear fields can be inferred from the variance of the matter density by the Poisson equation, $\Delta\Phi = 3\Omega_\mathrm{m}/(2\chi_H^2)\delta$, with the matter density $\Omega_m$ and the Hubble-distance $\chi_H=c/H_0$. Computing the tidal shear fields $\partial^2\Phi$ shows that they must have the same fluctuation statistics as the density field $\delta$. In Fourier-space, the solution to the Poisson equation is $\Phi \propto \delta/k^2$ with the wave vector $k$, and the tidal shear fields would become $k_ik_j\Phi \propto k_ik_j/k^2\:\delta$. Therefore, the spectrum $P_{\partial^2\Phi}(k)$ of the tidal shear fields is proportional to the spectrum $P_\delta(k)$ of the density fluctuations.

Consequently, one can derive the variance of tidal shear fields from the variance of the density fluctuations, i.e.\ from the CDM-spectrum $P_\delta(k)$. For doing that, one can relate a mass scale $M$ to the wave vector $k$ by requiring that the mass $M$ should be contained in a sphere of radius $R$, $M=4\pi\Omega_\mathrm{m}\rho_\mathrm{crit}R^3/3$, which in turn defines a scale $k$ in the spectrum which is proportional to $M^{-1/3}$. This implies that on galaxy- and cluster-scales, where the CDM-spectrum scales $\propto k^{-2\ldots-3}$, one obtains for the standard deviation of the tidal shear field a behaviour $\propto M^{1/2\ldots 1/3}$. We compare these very general theoretical predictions with our results in Sect.~\ref{sec:resanddiscuss}.
\section{Data}
\label{sec:data}
\subsection{Simulations}
\label{subsec:sim}
In this work we consider haloes from two different simulations: 
\begin{enumerate}
\item the \textbf{Millennium Simulation (MS)}, first presented in \citet{Springeletal2005}, which uses $2160^3$ dark matter particles of mass $m_{\mathrm{P}}^{\mathrm{MS}} =1.2 \times 10^9 M_{\sun}$ enclosed in a 500 Mpc$/h$-side box to sample dark matter haloes and study their growth. In particular, we take into consideration 2 of its 64 snapshots, i.e.\ the one at $z = 0$ (snapshot $63$), which we use for all our analysis, and the one at $z \simeq 0.46$ (snapshot $49$), which is studied only as an aside. Dark matter haloes are identified as in \citet{Joachimietal2013a} and references therein, assuming that all galaxies share the three-dimensional shape of their host halo. All haloes have a minimum number of particles $N_{\mathrm{P}} = 300$.
\item The \textbf{Millennium-XXL Simulation (XXL)}, which samples $6720^3$ dark matter particles of mass $m_{\mathrm{P}}^{\mathrm{XXL}} = 8.456 \times 10^9 M_{\sun}$ confined in a cubic region of $3000$ Mpc$/h$ on a side \citep{Anguloetal2012}; in this case, we consider only one snapshot, at $z = 0$. Haloes are selected using an ellipsoidal overdensity algorithm, as described in \citet{Despalietal2013} and \citet{Bonamigoetal2015}, assuming a minimum number of particles $N_{\mathrm{P}} = 1000$.
\end{enumerate} 
%the shape of the haloes is then derived as for the MS from the eigenvectors and eigenvalues of the mass tensor 
%\begin{equation}
%    \mathbf{M}_{\mu \nu} = m_P \sum_{i=1}^{N_P} \frac{r_{i, \mu} r_{i, \nu}}{M_{TOT}} = \frac{1}{N_P} \sum_{i=1}^{N_P} r_{i, \mu} r_{i, \nu}
%	\label{eq:mt}
%\end{equation}
%with $M_{TOT}$ the total mass of the object. Note that the mass tensor (Eq.~\ref{eq:mt}) and the inertia tensor (Eq.~\ref{eq:sit}) are different quantities \citep{Bettetal2007}, but the analysis on the two simulations yields pretty consistent results, as will be shown; also, for this catalogue no ``reduced'' tensor is taken into consideration.

@@

To compare our results with real data, we also take into consideration the analysis presented in \citet{vanUitertJoachimi2017}: in that work, the clusters contained in the redMaPPer catalogue \citep{Rykoffetal2014} version 6.3 were used to constrain the intrinsic alignment signal amplitude $A_{\mathrm{IA}}$, which was then studied as a function of the halo mass.

\subsection{Halo shapes}
\label{subsec:haloshapes}
We define the simple inertia tensor\footnote{MS and XXL use two different tensors to describe the shape, but our choice has no impact on the ellipticity of the haloes (see also \citet{Bettetal2007} for further details).}, whose eigenvalues and eigenvectors describe the shape of the halo, as:
\begin{equation}
    \mathbf{M}_{\mu \nu} \propto \sum_{i=1}^{N_{\mathrm{P}}} r_{i, \mu} r_{i, \nu} \ ,
	\label{eq:sit}
\end{equation}
where $N_{\mathrm{P}}$ is the total number of particles within the halo, and $\mathbf{r}_{i}$ is the vector that indicates the position of the $i-$th particle with respect to the centre of the halo, i.e.\ the location of the gravitational potential minimum. For the MS only, we also consider a reduced inertia tensor, which is defined as \citep{Pereiraetal2008}:
\begin{equation}
    \mathbf{M}_{\mu \nu}^{\mathrm{red}} \propto \sum _{i=1}^{N_{\mathrm{P}}} \frac{r_{i, \mu} r_{i, \nu}}{r_i^2} \ ,
	\label{eq:rit}
\end{equation}
with $r_i^2$ the square of the three-dimensional distance of the i-th particle from the centre of the halo; the reduced inertia tensor is more weighted towards the centre of the halo, and may yield a more reliable approximation of the shape of the galaxy \citep{Joachimietal2013b, Chisarietal2015}.

The eigenvectors define an ellipsoid, which we project onto one of the faces of the cube that hosts the simulation: the resulting ellipse is the shape of the galaxy, which is assumed to be the same as the shape of the host halo. We proceed as in \citet{Joachimietal2013a, Joachimietal2013b} to define the ellipticity $\epsilon$ of the galaxy, considering our objects as early-type galaxies: indicating the three unit eigenvectors as $\mathbf{s}_{\mu} = \big \{ s_{x, \mu}, s_{y, \mu}, s_{\mathbin{\|}, \mu}\big \}^\intercal$ and the absolute values of the semi-axes as $\omega_{\mu}$, $\mu \in \{1,2,3\}$, we define a symmetric tensor
\begin{equation}
    \mathbf{W}^{-1} = \sum _{\mu=1}^{3} \frac{s_{\perp, \mu} s^\intercal_{\perp, \mu}}{\omega_{\mu}^2} - \frac{\mathbf{\kappa}\mathbf{\kappa}^{\intercal}}{\alpha^2} \ ,
	\label{eq:symtensor}
\end{equation}
with $s_{\perp, \mu} = \big \{ s_{x, \mu}, s_{y, \mu}\big \}^\intercal$ the eigenvector projected along the line of sight, 
\begin{equation}
   \mathbf{\kappa} = \sum_{\mu = 1}^{3} \frac{s_{\mathbin{\|}, \mu} \mathbf{s}_{\perp, \mu}}{\omega_{\mu}^2} \ ,
	\label{eq:kappa}
\end{equation}
and 
\begin{equation}
    \alpha^2=\sum_{\mu = 1}^{3} \bigg( \frac{s_{\mathbin{\|}, \mu}}{\omega_{\mu}} \bigg)^2 \ .
	\label{eq:alpha}
\end{equation}
We then compute the two components of the ellipticity \citep{BartelmannSchneider2001}:
\begin{align}
    	\epsilon_{1} = \frac{W_{11} - W_{22}}{W_{11} + W_{22}+2\sqrt{\det \mathbf{W}}}\ ; \\ 
           \epsilon_{2} = \frac{2 W_{12}}{W_{11} + W_{22}+2\sqrt{\det \mathbf{W}}} \ .
\end{align}

@@

We describe the clustering signal with a simple model:
\begin{equation}
    w_{\mathrm{gg}} (R_{\mathrm{p}}, M)=b_g^2 (M)\ w_{\mathrm{\delta} \mathrm{\delta}}^{\mathrm{model}} (R_{\mathrm{p}}) - C_{\mathrm{IC}} \ , 
	\label{eq:wgg}
\end{equation}
with $w_{\mathrm{\delta} \mathrm{\delta}}^{\mathrm{model}} (R_{\mathrm{p}})$ a function in which we include the dependence on $R_{\mathrm{p}}$ \citep[equation 9]{vanUitertJoachimi2017}, and $C_{\mathrm{IC}}$ the integral constraint factor, which we estimate as in \citet[equation 8]{RocheEales1999} using the random pair counts. Again, we evaluate the previous expression in $R_{\mathrm{p}}^*$, obtaining:
\begin{equation}
    w_{\mathrm{gg}} (M)= w_{\mathrm{gg}} (R_{\mathrm{p}} = R_{\mathrm{p}}^*, M) \ .
	\label{eq:wgpmass}
\end{equation}

@@

To achieve this goal, we select the objects of the catalogues described in Sect.~\ref{subsec:sim} in $n = 4$ logarithmic mass bins, between $10^{11.5} M_{\sun}$ and $10^{14} M_{\sun}$ for the MS and between $10^{13} M_{\sun}/h$ and $10^{15} M_{\sun}/h$ for the XXL, we split them in $N = 3^3 = 27$ sub-boxes based on their positions inside the cube of the respective simulation, and calculate $w_{\mathrm{g+}}$ and $w_{\mathrm{gg}}$ for each of the N sub-samples by replacing the integrals in Eq.~\ref{eq:wgphat} and \ref{eq:wgghat} with a sum over 20 line-of-sight bins, each $(\Pi_{\mathrm{max}} - (-\Pi_{\mathrm{max}}))/ 20 \ = 6 \ \mbox{Mpc}/h$ wide. We define the mass of these objects as the mass within a halo which has mean density 200 times the critical value at the redshift corresponding to the respective snapshot $(M_{200\mathrm{c}})$; note that for the MS we first convert the halo mass from $M_{\mathrm{Dhalo}}$, as defined in \citet{Jiangetal2014}, to $M_{200\mathrm{c}}$ using the median line in \citet[figure 2]{Jiangetal2014}. We show the distribution of the selected masses for the two catalogues in Fig.~\ref{fig:histo}. 
\begin{figure}
	\centerline{
	%\includegraphics[height = 8 cm, width = 9 cm]{img/histo.eps}}
	\includegraphics[scale = 0.305, keepaspectratio]{img/histo.eps}}
	\caption{Histogram showing the density distribution of the mass of the haloes.
	% the Millennium-XXL catalogue (black line) contains about 6 times more objects than the Millennium catalogue (blue line). 
	In this case, the mass of the halo is defined considering an overdensity of 200 times the critical value $(M_{200\mathrm{c}})$.}
	% Note that in our analysis the Millennium-XXL catalogue is complete only above $10^{13} M_{\sun} h^{-1}$: below this threshold randomly selected haloes were used.}
	\label{fig:histo}
\end{figure}

To measure $DR$ and $RR$ as in Eq.~\ref{eq:xigghat}, we generate random catalogues that contain objects uniformly distributed between the minimum and maximum value of the $x$, $y$ and $z$ coordinates of each sub-box. These catalogues normally are 3 times denser; in some cases, when a sub-box encloses very few objects, we switch to random catalogues which are 10 times denser.

We then perform a posterior analysis over the data to retrieve the distributions of $A$ and $\beta$: according to Bayes' theorem, if $\boldsymbol{d}$ is the vector of the data and $\boldsymbol{p}$ the vector of the parameters, 
\begin{equation}
    P(\boldsymbol{p} | \boldsymbol{d}) \propto P(\boldsymbol{d} | \boldsymbol{p}) \ P(\boldsymbol{p}) \propto e^{-\frac{1}{2} \chi ^2} P(\boldsymbol{p}) \ ,
	\label{eq:bayes}
\end{equation}
with $P(\boldsymbol{p} | \boldsymbol{d})$ the posterior probability, $P(\boldsymbol{d} | \boldsymbol{p})$ the likelihood function, $P(\boldsymbol{p})$ the prior probability and $\chi ^2 = (\boldsymbol{d} - \boldsymbol{m})^T \mathbf{C}^{-1} (\boldsymbol{d} - \boldsymbol{m})$, with $\boldsymbol{m}$ the vector of the model and $\mathbf{C}^{-1}$ the precision matrix, the inverse of the covariance matrix $\mathbf{C}$. We assume uninformative flat priors in the fit with ranges $\log_{10} A \in [-2.3;-0.4]$ and $\beta \in [-0.50;0.75]$.

We estimate the covariance matrix from the data as in \citet{Tayloretal2013}:
\begin{equation}
     \mathbf{C}_{\mu \nu} = \frac{1}{N-1} \sum_{j = 1}^{N} (x_{j, \mu} - \overline{x}_{\mu})(x_{j, \nu} - \overline{x}_{\nu}) \ ,
	\label{eq:covariance}
\end{equation}
with $\mu, \nu \in \{1, \dotso, n\}, \overline{x}_{\mu} = \frac{1}{N} \sum_{j=1}^{N} x_{j, \mu},$ and $x_{j, \mu}= r_{\mathrm{g+}}(M)$ for each sub-box and each mass bin, as defined in Eq.~\ref{eq:rg+}. We then invert the covariance matrix and correct the bias on the inverse to obtain an unbiased estimate of the precision matrix, given by:
\begin{equation}
     \mathbf{C}^{-1}_{\mathrm{unbiased}} = \frac{N - n -2}{N-1} \  \mathbf{C}^{-1} \ ,
	\label{eq:precunbiased}
\end{equation}
where $N > n+2$ clearly holds \citep{Tayloretal2013}. The results of the analysis are presented in Sect.~\ref{sec:resanddiscuss}.

@@

%We thank Raul Angulo for helpful information about the Millennium-XXL catalogue snapshots, and Edo van Uitert for making the real data points available to us and for useful discussion about the correct mass units of measurement to use. BJ acknowledges support by an STFC Ernest Rutherford Fellowship, grant reference ST/J004421/1.

@@

\bibliographystyle{mnras}
\bibliography{biblio} % if your bibtex file is called example.bib

% Alternatively you could enter them by hand, like this:
% This method is tedious and prone to error if you have lots of references

%\begin{thebibliography}{99}

%\bibitem[\protect\citeauthoryear{Angulo et al.}{2012}]{Anguloetal2012}
%Angulo R.~E., Springel V., White S.~D.~M., et al., 2012, \href{https://academic.oup.com/mnras/article-lookup/doi/10.1111/j.1365-2966.2012.21830.x}{MNRAS}, \href{http://adsabs.harvard.edu/abs/2012MNRAS.426.2046A}{426}, \href{http://adsabs.harvard.edu/abs/2012MNRAS.426.2046A}{2046}
%\bibitem[\protect\citeauthoryear{Bartelmann \& Schneider}{2001}]{BartelmannSchneider2001}
%Bartelmann M., Schneider P., 2001, \href{http://www.sciencedirect.com/science/article/pii/S037015730000082X}{Phys.~Reports}, \href{http://adsabs.harvard.edu/abs/2001PhR...340..291B}{340}, \href{http://adsabs.harvard.edu/abs/2001PhR...340..291B}{291}
%\bibitem[\protect\citeauthoryear{Bett et al.}{2007}]{Bettetal2007}
%Bett P., Eke V., Frenk C.~S., et al., 2007, \href{https://academic.oup.com/mnras/article-lookup/doi/10.1111/j.1365-2966.2007.11432.x}{MNRAS}, \href{http://adsabs.harvard.edu/abs/2007MNRAS.376..215B}{376}, \href{http://adsabs.harvard.edu/abs/2007MNRAS.376..215B}{215}
%\bibitem[\protect\citeauthoryear{Bonamigo et al.}{2015}]{Bonamigoetal2015}
